A Transactional Memory System for Parallel Python

### 2014-07-30 talk/dls2014/paper/paper.tex
some corrections and additions

  Described below, we also need a write barrier for our generational
  garbage collector (GC). Interestingly, we are able to share the same
  barrier and flag for both use cases. Since the GC is an integral
  part of many dynamic languages, we actually do not incur any additional
  overhead here since the barriers are already there.

@@

Note that we have three kinds of objects: \emph{young, overflow} and
\emph{old} objects. Young and overflow objects were created in the
same transaction; old objects always come from previous transactions.
The close integration between STM and the GC allows us to use this
information effectively: young objects do not trigger the slow path at
all, overflow objects only need to be handled by the GC part of the
write barrier once between collections, and only the old objects need
to be handled by the STM part \emph{once per transaction}. This keeps the
overhead of the write barrier very low.

@@

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately not received much attention.
Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
  global interpreter locks in ruby through hardware transactional
  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '14).

>>>>

  Described below, we also need a write barrier for our generational
  garbage collector (GC). Interestingly, we are able to share the same
  barrier and flag for both use cases. Since the GC is an integral
  part of many dynamic languages, we often do not incur additional
  overhead since the barriers are already there. However, we do need
  some additional barriers when writing non-references into objects.
  In that case, generational GCs do not need a barrier but the STM
  system still does.

@@

Note that we have three kinds of objects: \emph{young, overflow} and
\emph{old} objects. Young and overflow objects were created in the
same transaction; old objects always come from previous transactions.
The close integration between STM and the GC allows us to use this
information effectively: young objects do not trigger the slow path at
all, overflow objects only need to be handled by the GC part of the
write barrier once between collections, and only the old objects need
to be handled by the STM part \emph{once per transaction}. The
privatisation step in particular is done very rarely because the pages
stay private until the next major collection where they may get
re-shared. This keeps the overhead of the write barrier very low in
the common cases.

@@

% STM needs good performance:
The solution presented here is based on a software TM system developed
as an independent and reusable C library for integration in dynamic
language interpreters. As the 
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately not received much attention.
Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
  global interpreter locks in Ruby through hardware transactional
  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '14).
